---
title: "Homework 1: simulate your own data"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```

In this homework problem, we will be building a `data.frame`. I will describe 
the final shape of the object and give you some hints on which functions to use.
However, as you've learned in the previous sections, there are ways to find more
about the functions you need, `?<function>` being the fastest one.

---

Here is what you need to do:

1. Create a `data.frame` with three columns and 30 rows.
2. The columns names should be `V1`, `V2` and `V3`.
3. `V1` will only be an ID variable that contains the numbers of the rows.
4. `V2` will be a random variable with mean 0, and a standard deviation of 1.
5. `V3` will repeat the letters 'A' and 'B', 15 times each.

You can check the functions `:`, `rnorm` and `rep`. Some of these we have used 
before, but some are new. The `?` is your friend. 

Below is an R console for you to write and run the code. Once you have it, you 
can move on to the questions below.

```{r practice,exercise=TRUE}
# create V1

# create V2
set.seed(123) # do not remove this line

# create V3

# create a data.frame
```

```{r practice-hint}
# create V1
V1 <- 1:30

# create V2
set.seed(123) # do not remove this line
V2 <- rnorm(30, mean = 0, sd = 1)

# create V3
V3 <- rep(c('A', 'B'), each = 15)

# create a data.frame d
d <- data.frame(V1, V2, V3)
```

If everything went fine, you should have a 30 by 3 `data.frame`. Check you
have it right using `str` in the console above. The questions will require you
to do subsetting as described before and perform some numerical operations. None
of the functions you need is new. You only need to figure out which pieces to 
use from the ones we have discussed in the practice.

```{r quiz}
quiz(
  question(
    text = "Which row in d has a V2 equals 1.71506499?",
    answer('6', correct = TRUE),
    answer('16'),
    answer('60')
  ),
  question(
    text = "What is the mean of the values in V2 when V3 == 'A'?",
    answer('0.15', correct = TRUE),
    answer('1.5'),
    answer('15'),
    random_answer_order = TRUE
  )
)
```
