---
title: "Homework 3: same genes, different data"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
library(pcr)
```

Now that you've learned to use `pcr` functions to model, visualize and test
$C_T$ values, you are ready to perform the full analysis on your own. 

The following is a description of the problem you need to solve. The data and
code are very similar to what you have seen in the practice session with one 
wrinkle. Read the description below and see if you can spot the difference. 
Once you do that, you will need to find a way to modify the code to accommodate
this difference and complete the analysis.

---

The goal of this exercise is to use `pcr_analyze` and `pcr_test` to calculate
the double delta $C_T$ value of the gene of interest in the `ct2` dataset. This 
dataset is very similar to `ct1`, however, they differ in one important aspect.
You will need to read the help pages of these datasets and figure out what this 
difference is and modify your code accordingly.

Here, are a few things to keep in mind

1. You need to load the `pcr` library to get access to the data and functions
2. To get the help page, use `?` followed by the name of the dataset or function
3. To modify the behavior of a function, check the relevant arguments
4. `pcr_analyze` is just an interface for other functions. The one relevant to 
this exercise is `pcr_ddct`

The following is an R prompt for you to work in.

```{r practice, exercise=TRUE}
# load pcr library


# inspect the data.frame ct2
# decide which is the reference gene

# compare the help pages of ct1 to ct2
# how the data were generated


# get the help page of pcr_ddct
# find the relevant argument to modify


# make a group variable for ct2
# decide which is the reference group


# call pcr_analyze


# call pcr_test


```

```{r practice-hint}
# load pcr library
library(pcr)

# inspect the data.frame ct2
# decide which is the reference gene
head(ct2, n = 3)

# compare the help pages of ct1 to ct2
# how the data were generated
?ct2
?ct1

# get the help page of pcr_ddct
# find the relevant argument to modify
?pcr_ddct

# make a group variable for ct2
# decide which is the reference group
group_var <- rep(c('brain', 'kidney'), each = 6)

# call pcr_analyze
pcr_analyze(ct2,
            group_var = group_var,
            reference_gene = 'GAPDH',
            reference_group = 'brain',
            mode = 'same_tube')

# call pcr_test
pcr_test(ct2,
         group_var = group_var,
         reference_gene = 'GAPDH',
         reference_group = 'brain')
```

Now based on your results, try to answer the following questions.

```{r quiz6}
quiz(
  caption = 'Practice',
  question(
    text = "What is the caliberated expression difference between kidney and brain?",
    answer("-2.53", correct = TRUE),
    answer("2.53"),
    answer("4.47"),
    answer("No idea")
  ),
  question(
    text = "What is the relative expression of MYC in kidney compared to brain tissue?",
    answer("1.0"),
    answer("6.99"),
    answer("5.77", correct = TRUE),
    answer("IDK")
  ),
  question(
    text = "What is the p-value after applying t-test to the expression difference?",
    answer("Above 0.00001"),
    answer("Below 0.00001", correct = TRUE),
    answer("Not sure")
  ),
  question(
    text = "What is the confidnece interval of the estimated expression difference?",
    answer("-1.76 to -1.29"),
    answer("-2.76 to -2.29", correct = TRUE),
    answer("-2.3 to -2.5"),
    answer("Who Knows?")
  )
)

```
