---
title: "Basic statistics in R"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE, fig.width = 5, fig.height = 5, fig.align = 'center')

# global variable
# create a grouping variable
set.seed(123)
d <- data.frame(
  V1 = 1:30,
  V2 = rnorm(30, mean = 0, sd = 1),
  V3 = rep(c('A', 'B'), each = 15)
)

d2 <- d

d2$V2 <- ifelse(d2$V3 == 'B', d2$V2 + 2, d2$V2)
```

## What is statistics? 

Statistics is the practice concerned with collecting and analyzing data. This
interactive practice is not strictly about statistics but rather, about doing
the basic statistical summarizing and testing in R.

We will only be looking at a few ideas in statistics that are common for every
data analysis. Namely

- Summarizing data
- Plotting data
- Applying statistical tests

---

We will start from where we ended in the previous module. The code below creates
three variables and collects them in a `data.frame`. You can think of this 
object, `d`, as encoding data from an experiment. In the following sense

- `V1`: the sample ID
- `V2`: measurements
- `V3`: a grouping variable

```{r df,echo=TRUE}
# create an index variable
V1 <- 1:30

# create a random variable
set.seed(123) # do not remove this line
V2 <- rnorm(30, mean = 0, sd = 1)

# create a grouping variable
V3 <- rep(c('A', 'B'), each = 15)

# create a data.frame d
d <- data.frame(V1, V2, V3)

# examine the object d
str(d)
```

Next, we will introduce some bias in the measurements and try to explore and 
retrieve it using statistical tools. Here is code that adds 1 to the 
measurements of group `'B'`, and leaves `'A'` as is.

```{r df2,echo=TRUE}
# copy object d to d2
d2 <- d

# add 1 to V2 when V3 == 'B'
d2[, 'V2'] <- ifelse(d[, 'V3'] == 'B',
                     d[, 'V2'] + 2,
                     d[, 'V2'])

str(d2)
```

Use `str`, `head` or `tail` to check whether the code above worked. In the next
section, we will use "summary statistics" to explore the bias we introduce 
between the groups A and B.

## Summary statistics 

Summarizing data is a sensible starting point when dealing with many data points
. Two useful summary statistics are averages and variance.

### Averages

Averages are single numbers that describe a list of numbers. There are more than
one way of defining it.

1. Mean: the sum divided by the length
2. Median: the value separating the upper half from the lower half
3. Mode: the value that appears most often

----

There are functions in R that correspond to the first two of these statistics. 
These are `mean` and `median`.

Here is the code to calculate the mean for all measurements and for each of the
two groups 'A' and 'B'.

```{r ave1,exercise=TRUE}
# calculate the mean of V2
mean(d2$V2)

# calculate the mean of the V2 for each group
mean(d2$V2[d2$V3 == 'A'])
mean(d2$V2[d2$V3 == 'B'])
```

Now try to calculate the other average statistic, `median`, for the same groups.

```{r ave2,exercise=TRUE}
# calculate the median of V2

# calculate the median of the V2 for each group

```

```{r ave2-solution}
# calculate the median of V2
median(d2$V2)

# calculate the median of the V2 for each group
median(d2$V2[d2$V3 == 'A'])
median(d2$V2[d2$V3 == 'B'])
```

Answer the following questions based on what you have done.

```{r quiz1}
quiz(
  question(
    text = "How does the median compare to the mean of V2?",
    answer("Median bigger than mean"),
    answer("Median smaller than mean", correct = TRUE),
    answer("Median equal to mean")
  ),
  question(
    text = "How does the median of group 'B' compare to group 'A'?",
    answer("Median bigger than mean", correct = TRUE),
    answer("Median smaller than mean"),
    answer("Median equal to mean")
  )
)
```

These single numbers tell us something about the center of the data points. 
Other statistics can tell us how much the data points vary from each other,
"Variance". We turn to this next.

### Variance

Variance is a measure of dispersion, meaning how far a set of numbers is spread 
out from their average value.

---

There functions `var` and `sd`, stand for variance and standard deviation.

Here is the code to calculate the variance and the standard deviation for all 
measurements.

```{r var1,exercise=TRUE}
# calculate the variance of V2
var(d2$V2)

# calculate the standard deviation of V2
sd(d2$V2)
```

Now try to calculate the same statistics for the groups 'A' and 'B' separately.

```{r var2,exercise=TRUE}
# calculate the variance and standard deviation for group A

# calculate the variance and standard deviation for group B

```

```{r var2-solution}
# calculate the variance and standard deviation for group A
var(d2$V2[d2$V3 == 'A'])
sd(d2$V2[d2$V3 == 'A'])

# calculate the variance and standard deviation for group B
var(d2$V2[d2$V3 == 'B'])
sd(d2$V2[d2$V3 == 'B'])
```

You should be able to answer the questions below based on the code above.

```{r quiz2}
quiz(
  question(
    text = "Which of the two groups has higher variance?",
    answer("A"),
    answer("B", correct = TRUE),
    answer("A and B has equal variance")
  ),
  question(
    text = "How does the variance relate to the standard deviation?",
    answer("Variance is the standard deviation squared", correct = TRUE),
    answer("Standard deviation is the variance squared"),
    answer("Variance is equal to the standard deviation")
  )
)
```

Point statistics, single number summaries, are very useful. But there are other 
ways to get a feel of how the data itself looks like. One way is to use graphs
to explore the data. We will be discussing some of the graphing ("Plots") 
tools in the next section.

## Plots 

There are many ways to visualize any given set of data. The basic graphs such as
bar, line and scatter plots are indispensable. Used in the right way and context
plots can be very illuminating. Once your data is large enough, graphs like 
these will be the only way to explore the data and the relationships between
the variables.

### Bar plot

Inspect the help page of the function `barplot` to see how to map the data to 
the bars.

```{r bar1,exercise=TRUE}
# inspect help page of barplot
```

The only required argument to `barplot` is called `height`, the rest is optional
(to make the plot look pretty). Here is an example

```{r bar2,echo=TRUE}
# plot 1 to 3
barplot(1:3)
```

Now, try plotting the `mean`s of the variable `d2$V2` for each group in `d2$V3`.
Remember, you can calculate the means just like before and save the results
in a variable of your choice, then pass this variable to `barplot`

```{r bar3,exercise=TRUE}
# calculate the group means

# plot the group means
```

```{r bar3-solution}
# calculate the group means
m <- c(
  mean(d2$V2[d2$V3 == 'A']),
  mean(d2$V2[d2$V3 == 'B'])
)

# plot the group means
barplot(m)
```

The exercise above made it look easy to plot numbers in R. Truth be told, it is
rarely the case that a single line of code would produce anything like a decent
graph. You'd have to modify it using other functions to make it informative.

The code below shows some embellishment to the above graph, just to show you how
it's done. But we won't be spending much time on it. I encourage you however to 
modify and execute the code to see how each function works.

```{r bar4,exercise=TRUE}
# calculate the group means
m <- c(
  mean(d2$V2[d2$V3 == 'A']),
  mean(d2$V2[d2$V3 == 'B'])
)

# plot the group means
barplot(m,
        xlab = 'Group',
        ylab = 'Measurement',
        main = 'Embellished Bar Plot',
        ylim = c(0, 2))
axis(side = 1,
     at = c(.7, 1.9),
     labels = c('A', 'B'))
box()
```

### Point plot

Another basic plot you can make is the point plot. The function `plot` creates
points with two coordinates 'x' and 'y'.

Here is code to plot all three variables in the object `d2`, which we created
earlier. The code puts `V1` on the x-axis, `V2` on the y-axis and colors the 
points by `V3`.

```{r point,echo=TRUE}
# plot all three variables
# as factor is necessary to force A and B to colors
plot(x = d2$V1,
     y = d2$V2,
     col = as.factor(d2$V3))
```

Modify the code below to change the labels of 'x' and 'y' to 'Variable 1' and 
'Variable 2'. You will need to pass these labels to the arguments `xlab` and 
`ylab`.

```{r point2,exercise=TRUE}
# plot all three variables
# as factor is necessary to force A and B to colors
plot(x = d2$V1,
     y = d2$V2,
     col = as.factor(d2$V3))
```

```{r point2-solution}
# plot all three variables
# as factor is necessary to force A and B to colors
plot(x = d2$V1,
     y = d2$V2,
     col = as.factor(d2$V3),
     xlab = 'Variable 1',
     ylab = 'Variable 2')
```

### Histogram plot

The third and last type of plot we will discuss is called a histogram. 
Histograms count the frequency of data points within specified ranges. In other 
words, it divides the data into ranges called 'bins' and counts how many data 
points occur in each bin.

The function `hist` can do that for you.

```{r hist,echo=TRUE}
# draw a histogram
hist(d2$V2[d2$V3 == 'A'],
     xlab = 'Range of variable 2',
     main = 'Histogram of V2')
```

Here is a question to test your understanding of histograms. One of these
can be answered directly for the histogram above, the second will require you
to make a tiny modification to the code.

```{r quiz3}
quiz(
  question(
    text = "What is the bin of the highest frequency in V2?",
    answer('-1 to 0'),
    answer('0 to 1', correct = TRUE),
    answer('1 to 2')
  ),
  question(
    text = "True or false: most of A values are above 0.5?",
    answer('True'),
    answer('False', correct = TRUE)
  )
)
```

## Tests 

Looking at the mean, variance, and plots, you can easily tell that 
the bias we introduced in the group 'B' is there. We introduced this 
difference intentionally. But in most cases, this would be the exact question we
ask of the data. Is there a difference between the measurements of the groups?

One way to answer this kind of question is to use statistical tests. 

---

Choosing which test to use, indeed whether you should use tests at all or 
whether they mean anything is not discussed.

Here is code that applies the a t-test to the two groups, 'A' and 'B'.

```{r test,exercise=TRUE}
# extract group values
A <- d2$V2[d2$V3 == 'A']
B <- d2$V2[d2$V3 == 'B']

# apply t.test
t.test(B, A)
```

Check the help page of the `t.test` function and try to interpret its output.

```{r quiz4}
quiz(
  question(
    text = "What is the estimated difference in mean A and B?",
    answer('4.49'),
    answer('1.6', correct = TRUE),
    answer('0.00013')
  ),
  question(
    text = "Is the difference between B and A significant at P < 0.001?",
    answer('Yes', correct = TRUE),
    answer('No'),
    answer('Not sure')
  )
)
```
