---
title: "Homework 4: two images at once "
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
library(colocr)
image1_file <- system.file('extdata', 'Image0001_.jpg', package = 'colocr')
image2_file <- system.file('extdata', 'Image0003_.jpg', package = 'colocr')
```

To process a collection of images at once, we only need to make tiny changes to
the workflow that we have been following.

- First, load multiple images in a list
- Second, pass the selection parameters as a list if they differ from one image 
to another.

---

Let's first start by loading the images in a single list. I located the files
in character vector and named it `img_files`. 

```{r img_files,echo=TRUE}
# locate image files
img_files <- c(image1_file, image2_file)
```

Use this object with `image_load` to create an `img_list`

```{r files,echo=TRUE}
# locate image files
img_files <- c(image1_file, image2_file)
```

```{r load5,exercise=TRUE,exercise.setup='files'}
# load two images at once
```

```{r load5-solution}
# load two images at once
img_list <- image_load(img_files)
```

---

Now we will select the regions of interest and test applying the same parameters
to both images. Only set `threshold = 90` and leave the rest with the default
values.

```{r imgs}
# load two images at once
img_files <- c(image1_file, image2_file)
img_list <- image_load(img_files)
```

```{r list,exercise=TRUE,exercise.setup="imgs"}
# call roi_select

# call roi_test
```

```{r list-solution}
# call roi_select
imgs_roi <- roi_select(img_list, threshold = 90)

# call roi_test
roi_test(imgs_roi)
```

You might however want to assign separate parameter values to each image. In
this case, you need to path a vector of the same length as the images instead of
a single value.

```{r list2,exercise=TRUE,exercise.setup="imgs"}
# declare thresholds for each image
threshold_list <- c(90, 85)

# call roi_select

# call roi_test
```

```{r list2-solution}
# declare thresholds for each image
threshold_list <- c(90, 85)

# call roi_select
imgs_roi <- roi_select(img_list, threshold = threshold_list)

# call roi_test
roi_test(imgs_roi)
```

Changing the threshold value in `roi_select` often results in different value of
the co-localization statistics. Therefore, you should select the parameters 
sensibly and often check how they change the selected regions. We discussed the 
`roi_show` which is designed for this purpose. 

---

Modify the code above and answer the following questions.

```{r quiz}
quiz(
  question(
    text = "What is the PCC value of 'image_2.jpg' at threshold 90?",
    answer("0.87"),
    answer("0.85", correct = TRUE),
    answer("0.83")
  ),
  question(
    text = "What is the PCC value of 'image_2.jpg' at threshold 85?",
    answer("0.87", correct = TRUE),
    answer("0.85"),
    answer("0.83")
  ),
  question(
    text = "What is the MOC value of 'image_2.jpg' at threshold 90?",
    answer("0.93"),
    answer("0.94", correct = TRUE),
    answer("0.95")
  )
)
```
