---
title: "Homework 2: plot your own data"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE, fig.width = 5, fig.height = 5, fig.align = 'center')

# create correlated variables
set.seed(1234)
N <- 50
V1 <- rnorm(50)
V2 <- rnorm(50, mean = .5 * V1)
d3 <- data.frame(V1, V2)
```

The dataset `d3` consists of two variables: `V1` and `V2`. Looking at the plot 
below, we can tell that the two variables are correlated, but it is hard to tell
by how much and whether the correlation is a significant one.

```{r df3}
plot(d3$V1, d3$V2,
     xlab = 'Variable 1',
     ylab = 'Variable 2',
     main = 'Correlation = ??',
     col = 'red',
     lwd = 2)
```

---

Here is what you need to do to answer the questions that follow.

1. Calculate the correlations between `V1` and `V2` using `cor`
2. Test whether these correlations are significant
3. Reproduce the plot above after replacing the question mark with the value of
correlation you calculated.

```{r practice,exercise=TRUE}
# show d3
str(d3)

# calculate correlation using cor

# test correlation using cor.test

# reproduce the plot using, well, plot
```

```{r practice-hint}
# calculate correlation using cor
cor(d3$V1, d3$V2)

# test correlation using cor.test
cor.test(d3$V1, d3$V2)

# reproduce the plot using, well, plot
plot(d3$V1, d3$V2,
     xlab = 'Variable 1',
     ylab = 'Variable 2',
     main = paste0('Correlation = ', round(cor(d3$V1, d3$V2), 2)),
     col = 'red',
     lwd = 2)
```

```{r quiz}
quiz(
  question(
    text = "What is the correlation coefficient between V1 and V2?",
    answer('0.24'),
    answer('0.34', correct = TRUE),
    answer('0.44')
  ),
  question(
    text = "What is the p-value of the correlation?",
    answer('0.013'),
    answer('0.014'),
    answer('0.015', correct = TRUE)
  )
)
```
