---
title: "Getting started in R"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```

## What is R? 

You can think of R as a fancy calculator. One that can perform operations on 
numbers and other types of data. 

There are a few reasons why learning to do data analysis in R can be useful to
you.

- Store, manipulate, analyze and visualize data
- Scale data analysis
- Benefit from open-source tools and methods
- Reproducible analysis

Notice however that most of these benefits can be accrued from other command-
line tools.

--- 

Once you get over the hurdle of expressing your experimental design in code, you
will come to appreciate how easy it is to do data analysis in R.

---

There are a few things you need to know to get started using R. These are things
called variables, functions, data types, data frames and subsetting. We will 
go through these one by one. Along the way you will learn about other things 
such as storing data and performing arithmetic.

But there is much more that you will need to learn once you know what to look 
for. That's why we will also discuss the different ways you can get help when 
stuck.

## Variables 

To drive home the calculator metaphor, look at the following line of code.

```{r var1,echo=TRUE}
1 + 2
```

The code above adds one plus two, just like a calculator. Take a look at 
the code below, and notice how it is different from the one before.

```{r var2,echo=TRUE}
x <- 1
x + 2
```

`x` here is a variable and it contains the value we assign to it using this
symbol `<-`. Adding `x` and `2` gives the same results as `1 + 2`. Using 
variables rather than typing the values  is necessary for reasons that will
become clear later. But here is a flavor of it. Type `x` in the box below and
inspect the results.

```{r x}
x <- 1
```

```{r var3,exercise=TRUE,exercise.setup='x'}
# type x and run the code
```

```{r quest-1}
question("What is the value of x?",
  answer('1', correct = TRUE),
  answer('"1"'),
  answer('x')
)
```

The value of `x` stays the same as long as you don't change it. So you can use
it over and over in your code. This will become very important once you have
many data points inside the variable. Which brings us to the next topic, 
"Functions".

## Functions 

Say you want to assign multiple values to `x`. For that you need a function
called `c`, it stands for combine. In fact, whenever you want to do anything in
R, you should be first looking at the function that does it. If there isn't one
you can create it yourself. Let's see how this works.

---

Here is code that stores the values `1`, `2`, and `3` in a variable called `x`
and prints it out.

```{r fun1,echo=TRUE}
# make a variable x
x <- c(1, 2, 3)
x
```

`x` is no longer a single value, but three. Thanks to the function `c` which
combines any number of values you put between parenthesis, `()`, separated by 
commas, `,`.

---

Let's take a look at two other functions. `sum` takes a variable and returns its
sum, `length` takes a variable and returns its length. Try them out.

```{r fun2, execise=TRUE}
# make a variable x
x <- c(1, 2, 3)

# call sum
sum(x)

# call length
length(x)
```

I think you can appreciate now why storing values in variables can be useful.
We achieved this through the function `c`. But functions can do much more.

Let's use one more. The function `mean` takes a variable and returns 
its mean (average). `x` is now assigned big numbers and its probably difficult
to guess the average, so try calling `mean` on `x`.

```{r fun3,exercise=TRUE}
# make a variable x
x <- c(154, 223, 309)

# call mean
```

```{r fun3-solution}
# make a variable x
x <- c(154, 223, 309)

# call mean
mean(x)
```

```{r quest-2}
question(
  text = "What is the mean of the variable x?",
  answer("686"),
  answer("223"),
  answer("228.67", correct = TRUE)
)
```

---

Imagine however that no such function as `mean` exists. Then you can think of a
way to compute your target using existing functions. Knowing that the mean of a
variable is the sum divided by the length. Can you think of a way to compute it?

Here is a hint: the output of a function is itself a variable and functions can
be combined in one line to produce other variables.

Remember

- `sum` calculates the sum of items in a variable
- `length` counts the number of items in a variable
- `/` is used for division just like `+` for addition

```{r fun4,exercise=TRUE}
# make a variable x
x <- c(154, 223, 309)

# compute the mean
```

```{r fun4-solution}
# make a variable x
x <- c(154, 223, 309)

# compute the mean
sum(x) / length(x)
```

Your output should match the one we got using `mean`. If so, you might want to 
save this way of computing the mean in a new function to be able to use it 
later. 

---

There is a function called `function` that is used to create functions. Just in
the last sentence, I have used the word "function" three times. You might be 
surprised to discover that it is much easier to express that thought in code.

```{r fun5,echo=TRUE}
# create a function to compute the mean
newfun <- function(x) sum(x) / length(x)

# call new fun
x <- c(154, 223, 309)
newfun(x)
```

Congratulations. You just created your first R function. This function `newfun`
takes as an input a variable `x` and computes its mean. It does this by dividing
the sum by the length of the variable.

```{r quest-3}
question(
  text = "What does x refer to in function(x)?",
  answer("The variable x we created before."),
  answer("Any variable we pass to newfun."),
  answer("Any variable of numbers we pass to newfun.", correct = TRUE)
)
```

The answer to this question takes us straight to the next section about "Data
types".

## Data types 

So far we have been working with variables that consist of numbers. But there
are other types of variables we can create. 

The most basic ones are

1. Numeric: real numbers such as 1.1, 1.2, or 1.3
2. Integer: whole numbers such as 1, 2, or 3
3. Character: strings of letters such as 'name', 'age', or 'gender'
4. Logical: either TRUE or FALSE

---

Let's create some variables and examine the differences between them.

```{r types1,echo=TRUE}
# create a numeric
num <- c(1.1, 1.2, 1.3)

# create an integer
int <- c(1L, 2L, 3L)

# create a character
char <- c('name', 'age', "gender")

# create a logical
logi <- c(TRUE, TRUE, FALSE)
```

```{r vars}
# create a numeric
num <- c(1.1, 1.2, 1.3)

# create an integer
int <- c(1L, 2L, 3L)

# create a character
char <- c('name', 'age', "gender")

# create a logical
logi <- c(TRUE, TRUE, FALSE)
```

Notice that although we can use the same function `c` to create all types, 
each has to be coded in a specific way. For example

- Integers are signified by `L` after the number.
- Characters are enclosed between single quotes `''` or double quotes `""`.
- Logicals are uppercase without quotes.

These variables also behave differently. You will come to appreciate them in
time since they make things easier when it comes to data analysis.

To confirm what type a variable is, use the function `class`. Here is an example

```{r types2,exercise=TRUE,exercise.setup='vars'}
# examine a variable class
class(num)
```

```{r quest-4}
question(
  text = "What is the class of the variables int, char, and logi?",
  answer("numeric, character, and logical."),
  answer("integer, character, and logical.", correct = TRUE),
  answer("factor, character, and logical.")
)
```

---

You can convert some data types into others. Often an `as.<class>` function 
exists to do just that. For example there are `as.integer`, `as.numeric`, 
`as.character` and `as.logical`. Try calling these functions on the variable 
`num` and examine the output.

```{r types3,exercise=TRUE,exercise.setup='vars'}
# call as.integer
as.integer(num)

# call as.character

# call as.logical

```

```{r types3-solution}
# call as.integer
as.integer(num)

# call as.character
as.character(num)

# call as.logical
as.logical(num)
```

```{r quest-5}
quiz(
  question(
    text = "What happens when you call as.integer(num)?",
    answer("Same number are returned."),
    answer("Integer part of the numbers are returned.", correct = TRUE)
  ),
  question(
    text = "What happens when you call as.character(num)?",
    answer("Same number are returned."),
    answer("Same number are returned as strings between quotes.", correct = TRUE)
  ),
  question(
    text = "What happens when you call as.logical(num)?",
    answer("Same number are returned."),
    answer("TRUE is repeated as many times as there are items in num.", correct = TRUE)
  )
)
```

You might be wondering why these variables behave the way they do. Mainly, each
of these types has a direct analogy to the real-world data.
It makes life easier for you and your computer, but at the cost of a bit of 
confusion like the one you are feeling right now.

---

Here is an example. What if we want to tell which of the values in `num` equals 
or differs from a number. There we can use operators such as `==`, `>` or `<`.
Although the inputs are all `numeric`, the output will be `logical`. Check the 
following code to see for yourself.

```{r types4,exercise=TRUE,exercise.setup='vars'}
# which value in num is equal to 1.2
num == 1.2

# which value in num is bigger than 1.2

# which value in num is smaller than 1.2
```

```{r types4-solution}
# which value in num is equal to 1.2
num == 1.2

# which value in num is bigger than 1.2
num > 1.2

# which value in num is smaller than 1.2
num < 1.2
```

Each of the variables (`vector`s) above had one type of data in them. This is a
simple and useful way of encoding data. However, most of the time data will come
in tables, and in mixed types: numbers, characters and all.

The next section will take you through one way to make tables with different 
types of data in them. These are called "Data frames".

## Data frames 

`data.frame`s are basically tables. They consist of rows and columns. Columns 
can have different types of data. While learning to do data analysis in R, 
`data.frame`s are probably the most important data structure and the most 
commonly used.

---

Let's see an example of how to build one from scratch.

```{r frame1,echo=TRUE}
# make variables
x <- rep(c(1, 2), each = 3)
x

y <- rep(c('A', 'B'), each = 3)
y

# make a data.frame
d <- data.frame(x, y)
d
```

```{r df}
x <- rep(c(1, 2), each = 3)
y <- rep(c('A', 'B'), each = 3)
d <- data.frame(x, y)
```

The `data.frame` we just created consists of two columns: `x` and `y`. We 
created `x` and `y` by repeating two values `c(1, 2)` and `c('A', 'B')`, 3 times each. The function `data.frame`, takes as input the columns and returns, well, a 
`data.frame`.

--

This object, `d`, is small, so you can print it out in its entirety without a 
problem. However this is not always the case. The function `str` shows you
the most important info you need about the object without having to print it out
on the screen.

```{r frame2,exercise=TRUE,exercise.setup='df'}
# call str on the object d
```

```{r frame2-solution}
# call str on the object d
str(d)
```

From the output above, you should be able to answer the following questions.

```{r quiz-6}
quiz(
  question(
  text = "How many rows and columns are there in the object d?",
  answer("6 and 2", correct = TRUE),
  answer("2 and 6"),
  answer("1 and 2"),
  random_answer_order = TRUE
),
question(
  text = "How many observations and variables are there in the object d?",
  answer("6 and 2", correct = TRUE),
  answer("2 and 6"),
  answer("1 and 2"),
  random_answer_order = TRUE
),
question(
  text = "How what are the data types of columns in the object d?",
  answer("Numeric and character", correct = TRUE),
  answer("Charcters only"),
  answer("Numerics only"),
  random_answer_order = TRUE
)
)
```

That's it about `data.frame`s for now. We will be using them again soon. But 
first, you need to learn how to break them down into pieces, otherwise known as
"subsetting", which is what we turn to next.

## Subsetting 

Just like there is a way to build bigger objects, there is a way to subset them.
Why would you need to do that? Well, you may want to have a look at a slice of
the object, filter out some unwanted data, or test your code on parts of the 
object before applying it to the whole dataset.

---

Here is the same code to generate the object `d` as before.

```{r frame3,echo=TRUE}
# make variables
x <- rep(c(1, 2), each = 3)
y <- rep(c('A', 'B'), each = 3)

# make a data.frame
d <- data.frame(x, y)

# show the structure of d
str(d)
```

Say we are only interested in the first column, `x`. Here, is a way to subset it
from `d` and create a new variable called `dx`. Run the code and call `str`
on the output.

```{r frame4,exercise=TRUE,exercise.setup='df'}
# subset first column
dx <- d[, 1]
str(dx)
```

Now do the same for the column `y` and call it `dy`.

```{r frame5,exercise=TRUE,exercise.setup='df'}
# subset second column
```

```{r frame5-solution}
# subset second column
dy <- d[, 2]
str(dy)
```

```{r quiz-7}
question(
  text = "What is the data type of dy?",
  answer("Character", correct = TRUE),
  answer("Numerics"),
  answer("data.frame"),
  random_answer_order = TRUE
)
```

---

To subset the first row of the object `d`, you do something similar.

```{r frame6,exercise=TRUE,exercise.setup='df'}
# subset the first row
d[1,]
```

In the code above, we used square brackets `[]` and inside them, the indecies of
the columns or rows we wanted.

---

You can subset a `data.frame` by column and rows at once. You can also get more
then one column or row at once. Just substitute the index between `[]` with a
vector of more then one index. Try this out.

```{r frame7,exercise=TRUE,exercise.setup='df'}
# subset the first three rows and two columns
```

```{r frame7-solution}
# subset the first three rows and two columns
d[c(1, 2, 3), c(1,2)]

# or
d[1:3, 1:2]
```

There are other ways to subset `data.frame`s. Indeed there is much more to learn
about them. One aspect of mastering a language like R is to be able to find help
on your own. This will be the topic of the next section, "Getting help".

## Getting help 

One of the benefits of using a statistical programming language is that you are
never alone. Pretty much any kind of analysis you would wish to do, someone else
has thought of. This also applies to learning the syntax of the language and 
solving the common issues that beginner face. 

These are ways to get help

- Type a question mark `?` followed by the name of a package, function, or class
- Copy and paste errors and warnings into your preferred search engine
- Trial and error
